A Computational Account of Default-Mode Function\\ by Control Theory and Reinforcement Learning

### 2016-10-01 paper/paper.tex
new comments from Silver video 1+2

\section{Reinforcement-Learning as a computational model for DMN function}
The main idea is to show that the DMN has biological substrates (gradient
computation, backprop, experience replay, etc.) for all the
major components of an efficient RL algorithm, and all these components
integrate in a plausible way.
\subsection{Markov Decision Processes}
In a general reinforcement-learning (RL) framework, an \textit{agent}
interacts with an  environment in a trial-by-error fashion by taking
actions / decisions $a$, which trigger changes in the
\textit{state} of the environment
$s \rightarrow s'$, accompanied with a real-valued \textit{reward}
$r = r(s, a)$ (or regret!, since it can be negative) collected by the
agent. This reward should be thought
of as an instantaneous satisfaction accompanying the execution of
an action. Here the subscript $t$ denotes time, and can be discrete or
continuous.
%% The environment $\mathcal E$ is generally taken as stochastic, but having some kind of
%% memory.
A popular abstract model for the scenario are so-called \textit{Markov Decision Processes (MDPs)}.
This is simply a quintuplet $(\mathcal S, \mathcal A, r, \rho, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, e.g $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $\mathcal A$ is the set of actions, e.g $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$, so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts-off in state $s$.
  \item $\rho : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto \rho(s'|s,a)$, the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such transitions be Markovian.
\end{itemize}

@@

Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s_t, a_t)$, a cumulative version which
takes into account the future. A standard for modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  R(s_t, a_t) = r(s_t, s_t) + \ldots + \gamma^Tr(s_{t+T}, s_{t+T})
  = r(s_t, s_t) + \gamma \sum_{\tau=1}^{T}\gamma^{\tau-1}r(s_{t+\tau},a_{t+\tau}),
  \end{equation}
for some \textit{discount factor} $\gamma \in [0, 1]$, and horizon $T$. The choice $\gamma = 0$ (or $T=0$)  corresponds to a perfectly myopic agent who is solely concerned about their immediate rewards and have no regards for the future. For there to be any learning at all, a technical requirement is that $\gamma < 1$ \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus the agent is positively biased towards considering rewards which are will
come soon. Note that $R(s_t, a_t)$ defined above is a stochastic quantity which depends on the trajectory generated by the policy $\pi$, upto but not including time $t$
\begin{eqnarray*}
  \begin{split}
    s_0 &\sim \mu\\
    a_0 &\sim \pi(.|s_0)\\
    s_1 &\sim \rho(.|s_0,a_0) \text{ and collect rewards }r(s_0, a_0)\\
    a_1 &\sim \pi(.|s_1)\\
    s_2 &\sim \rho(.|s_1,a_1) \text{ and collect rewards }r(s_1, a_1)\\
    \ldots\\
    a_{t-1} &\sim \pi(.|s_{t-1})\\
    s_{t} &\sim \rho(.|s_{t-1},a_{t-1}) \text{ and collect rewards }r(s_{t-1}, a_{t-1})
  \end{split}
\end{eqnarray*}

@@

\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The celebrated \textit{Bellman's equation} [Sutton and Barto 1998] then takes the simple form
\begin{equation}
  Q^\pi(s_t, a_t) = \mathbb E_{s_{t+1} \sim \rho(.|s_t,a_t)} [r(s_t,a_t) + \gamma Q^\pi(s_{t+1}, \pi(s_{t+1}))]
  \label{eq:bellman}
\end{equation}
which is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

\subsection{Efficient RL with value approximation and replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
We need ways to approximate the state-action value function
which scale up to large state spaces and / or action spaces.
In state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy matrix, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
and the environment transitions to a new state $s' \in \mathcal S$. At the end of
this step, a new \textit{experience} $e = (s,a,r,s')$ is produced; this represents an exemplar
behavior of the agent and is recorded in replay memory buffer:
$\mathcal D \leftarrow \mathcal D \cup \{e\}$ (possibly discarding the oldest entries to make space).
Now, at iteration $k+1$, replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$ and trying to
approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by Bellman's equation \eqref{eq:bellman}, namely
$r + \gamma \max_{a'} Q(s', a'|\theta^Q_k)$, with the output of a parametrized regression model $(s,a,r,s')
\mapsto {Q}(s, a|\theta^Q_{k})$, where $Q(.,.|\theta^Q_{k})$ is an approximator for the $Q$-value operator,  parameterized by $\theta^Q_{k+1}$.
Computing an optimal policy then corresponds to finding the parameters $\theta^Q_{k+1}$ which minimize the following mean-squared loss function
\begin{equation}
  \mathcal L(\theta^Q_{k+1})
  % = \mathbb E_{(s, a, r, s') \sim U(\mathcal D)}[(Q(s, a|\theta^Q) - Q(s',a'|\theta^Q_{k+1}))^2]
  = \mathbb E_{(s, a, r, s') \sim U(\mathcal D)}\left[\frac{1}{2}(Q(s, a|\theta^Q_{k+1}) - y)^2\right],
  \label{eq:oracle}
\end{equation}
where
$ y = r + \gamma \max_{a' \in \mathcal A} Q(s', a'|\theta^Q_k) = r + \gamma Q(s', \pi(s')|\theta^Q_k)$ is the prediction target.
% While $y_t$ is also dependent on $\theta^Q$, this is typically ignored.
For example, a general linear model with a kernel $\phi$ would be of the
form
$${Q}(s, a|\theta^Q) = \phi(s,a)^T\theta^Q.$$
$\phi(s,a)$ would represent a high-level hand-crafted representation of the state-action $(s,a)$ into an appropriate
feature space. Such are the techniques that have been proposed by [sutton et al. xyzw, Song et al. 2016 (NIPS),
  and Parr et al. 2003 (JMLR)].
A breakthrough alternative proposed by Google DeepMind [Minh et al. 2014, Silver et al 2015] is to learn this
representation using a deep neural net, leading to the so-called \textit{Deep Q-learning} family of methods which
are by far the current state-of-the-art in RL. In such a case, $\theta^Q$ would correspond to ..

@@

\paragraph{Parameter update.}
\begin{equation}
  \nabla_{\theta^Q_{k+1}}\mathcal L(\theta^Q_{k+1})
  % = \mathbb E_{(s, a, r, s') \sim U(\mathcal D)}[(Q(s, a|\theta^Q) - Q(s',a'|\theta^Q_{k+1}))^2]
  = \mathbb E_{(s, a, r, s') \sim U(\mathcal D)}\left[\underbrace{(Q(s, a|\theta^Q_{k+1}) - y)}_{\text{regret}}
    \underbrace{\nabla_{\theta^Q_{k+1}}Q(s, a|\theta^Q_{k+1})}_{\text{???}}\right],
  \label{eq:oracle}
\end{equation}

\paragraph{Link to classical RL algorithms.}
Note that most RL algorithms, including \textit{Temporal Difference (TD)} learning [Sutton et al.,
  Bertsekas] and SARSA can be cast in this general variational framework. For example TD can corresponds
to the above framework with a linear value approximator with feature encoding $\phi(s,a) = \delta_{(s,a)} =  \text{ point mass at }(s,a)$,
and so $$\nabla_{\theta^Q_{k+1}}Q(s, a|\theta^Q_{k+1}) = \phi(s, a) = \delta_{(s,a)}$$
...

>>>>

\section{Reinforcement-Learning as a computational model for DMN function}
The main idea is to show that the DMN has biological substrates (gradient
computation, backprop, experience replay, etc.) for all the
major components of an efficient RL algorithm, and all these components
integrate in a plausible way.
\subsection{Markov Decision Processes}
In a general reinforcement-learning (RL) framework, an \textit{agent}
interacts with an  environment in a trial-by-error fashion by taking
actions / decisions $a$, which trigger changes in the
\textit{state} of the environment
$s \rightarrow s'$, accompanied with a real-valued \textit{reward}
$r = r(s, a)$ (or regret!, since it can be negative) collected by the
agent. This reward should be thought
of as an instantaneous satisfaction accompanying the execution of
an action. Here the subscript $t$ denotes time, and can be discrete or
continuous.
The environment $\mathcal E$ is generally taken as stochastic,
is partially observed and 
has some kind of memory.
A popular abstract model for the scenario are so-called \textit{Markov Decision Processes (MDPs)}.
In RL, MDPs are the mathematical instantiation of the environment model,
which generally can incorporate action probabilities (policy matrix)
and/or value function given a state.
(Silver: The environment can be fully defined by state space and
transition probability matrix.)
This is simply a quintuplet $(\mathcal S, \mathcal A, r, \rho, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is any function of the history.
The state exhibits the Markov property
$\mathbb{P}(S_{t+1}|S_{t})=\mathbb{P}(S_{t+1}|S_{1}, \dots, S_{t})$.
Consequently, the future states are independent of past states and only depend on the present.
% \item $\mathcal S$ is the set of states, e.g $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $\mathcal A$ is the set of actions, e.g $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$, so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts-off in state $s$.
  \item $\rho : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto \rho(s'|s,a)$, the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such transitions be Markovian.
\end{itemize}

@@

Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s_t, a_t)$, a cumulative version which
takes into account the future. A standard for modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  R(s_t, a_t) = r(s_t, s_t) + \ldots + \gamma^Tr(s_{t+T}, s_{t+T})
  = r(s_t, s_t) + \gamma \sum_{\tau=1}^{T}\gamma^{\tau-1}r(s_{t+\tau},a_{t+\tau}),
  \end{equation}
for some \textit{discount factor} $\gamma \in [0, 1]$, and horizon $T$. The choice $\gamma = 0$ (or $T=0$)  corresponds to a perfectly myopic agent who is solely concerned about their immediate rewards and have no regards for the future. For there to be any learning at all, a technical requirement is that $\gamma < 1$ \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus the agent is positively biased towards considering rewards which are will
come soon. The reward importance hence decays with distance in the future,
given that there is more uncertainty in the farsighted future.
The reward discount $\gamma$ calibrates the delay discounting
behavior of the intelligent agent.

Note that $R(s_t, a_t)$ defined above is a stochastic quantity which depends on the trajectory generated by the policy $\pi$, upto but not including time $t$
\begin{eqnarray*}
  \begin{split}
    s_0 &\sim \mu\\
    a_0 &\sim \pi(.|s_0)\\
    s_1 &\sim \rho(.|s_0,a_0) \text{ and collect rewards }r(s_0, a_0)\\
    a_1 &\sim \pi(.|s_1)\\
    s_2 &\sim \rho(.|s_1,a_1) \text{ and collect rewards }r(s_1, a_1)\\
    \ldots\\
    a_{t-1} &\sim \pi(.|s_{t-1})\\
    s_{t} &\sim \rho(.|s_{t-1},a_{t-1}) \text{ and collect rewards }r(s_{t-1}, a_{t-1})
  \end{split}
\end{eqnarray*}

@@

\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The celebrated \textit{Bellman's equation} [Sutton and Barto 1998] then takes the simple form
\begin{equation}
  Q^\pi(s_t, a_t) = \mathbb E_{s_{t+1} \sim \rho(.|s_t)} [r(s_t,a_t) + \gamma Q^\pi(s_{t+1}, \pi(s_{t+1}))]
  \label{eq:bellman}
\end{equation}
which is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

\subsection{Efficient RL with value approximation and replay}
As already mentioned in the previous section, Q-learning optimizes over the class of
deterministic policies of the form \eqref{eq:qlearning}.
in state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy matrix, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
and the environment transitions to a new $s' \in \mathcal S$. At the end of
this step, a new \textit{experience} $e = (s,a,r,s')$ is produced; this represents an exemplar behavior of the agent and is recorded in replay memory:
$\mathcal D \leftarrow \mathcal D \cup \{e\}$.
Now, replay consists in sampling (uniform or importance-weighted\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$ and trying to
approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by Bellman's equation \eqref{eq:bellman}, namely
$r + \gamma \max_{a'} Q(s', a'|\theta^Q)$, with the output of a parametrized regression model $(s,a,r,s') \mapsto {Q}(s, a|\theta^Q)$, where $Q(.,.|\theta^Q)$ is an approximator for the $Q$-value operator,  parameterized by $\theta^Q$. Computing an optimal policy then corresponds to finding the parameters $\theta^Q$ which minimize the following loss function
\begin{equation}
  \mathcal L(\theta^Q)
  % = \mathbb E_{(s, a, r, s') \sim U(\mathcal D)}[(Q(s, a|\theta^Q) - Q(s',a'|\theta^Q))^2]
  = \mathbb E_{(s, a, r, s') \sim U(\mathcal D)}[(Q(s, a|\theta^Q) - y)^2],
  \label{eq:oracle}
  \end{equation}
where
$ y = r + \gamma \max_{a' \in \mathcal A} Q(s', a'|\theta^Q) = r + \gamma Q(s', \pi(s')|\theta^Q)$ is the prediction target.
% While $y_t$ is also dependent on $\theta^Q$, this is typically ignored.

For example, a general regression model with a kernel $(s, a) \mapsto \phi(s, a)$ would be of the
form

$${Q}(s, a|\theta^Q) = \phi(s,a)^T\theta^Q.$$
$\phi$ would represent a high-level hand-crafted representation of the state-action pairs into an appropriate feature space. Such are the techniques that have been proposed by [sutton et al. xyzw, Song et al. 2016 (NIPS), and Parr et al. 2003 (JMLR)].
A breakthrough alternative proposed by Google DeepMind [Minh et al. 2014, Silver et al 2015] is to learn this representation using a deep neural net, leading to the so-called \textit{Deep Q-learning} family of methods which are by far the current state-of-the-art in RL. In such a case, $\theta^Q$ would correspond to ..

@@

\begin{remark}
Note that most RL algorithms, including \textit{Temporal Difference (TD)} learning [Sutton et al.,
  Bertsekas] and SARSA can be cast in this general variational framework.
\end{remark}
